Learning how to act: making good decisions with machine learning

### 2017-04-30 thesis/thesis_map.tex
bootcamp sprint3

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{natbib}
\usepackage{wrapfig}

@@

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

@@

@@

@@

%For image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. In fact this is because we already have one. We can be confident that the process will not change. If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained). Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable. The mechanism is known but the function may be complex.

\subsubsection*{What aspects of a problem determine if causal inference is required?}
(When is pure prediction useful?)
\begin{itemize}
\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and priortise those for modelling changes. 
\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its ok...
\end{itemize}

@@

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}). If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network. 
%
%\begin{figure}[h]
%\centering
%\caption{General Bayesian network over 3 variables}
%\label{fig:bayesnet}
%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
%  thick,main node/]
%
% %nodes
%\node[main node](1){$X_{1}$};
%\node[main node, below left=of 1](2){$X_{2}$};
%\node[main node, below right=of 1](3){$X_{3}$};
%
%
% \path[every node/.style={font=\sffamily\small}]
%    (1) edge node {} (2)
%    	edge node {} (3)
%    (2) edge node {} (3);
%	
%\end{tikzpicture}
%\end{figure}

@@

A causal network represents much more information than a bayesian network with identical structure. GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS. This may be surprising since their conditional probability distributions are estiated from data in the same way. It is only possible due to the additional assumptions inherent in a causal network.

@@

\subsubsection*{Limitations of causal bayesian networks}
A number of critisims have been leveled at this way of modelling causality.

A complaint leveled against this view point of causality is that the 'surgery' is too precise and that, in the real world, any intervention will effect many variables (eg Cartwright 2007). However,

What questions can't be phrased using this terminology
- certain counterfactual queries
- cyclic/dynamics

Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.

- add variables for soft interventions
- unroll to deal with cycles

@@

There is a problem with the term non-parametric. It is used in machine learning in a slightly different way to pearl

@@

Once the data is broken down by gender, it appears that treament B is better for both men \emph{and} women. This is the paradox. If the doctor were unaware of the gender, it seems that she should advise treatment A. But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be. How does causaul modelling reslove this paradox? Let's represent the treatment presribed by the variable $T$, the gender of the patient by $Z$ and wheter or not the treatment was successful by $Y$. The key to resolving the paradox is that the doctor really cares about the interventional distribution $P{Y|do(T)}$, not the standard conditional distributions $P{Y|T}$. The results from table XXX shows that $\P{Y|T=A}$ is higher than $\P{Y|T=B}$. If she has to suggest a treatment for a patient without knowing their gender, she should choose the treatment for which $P{Y|do(T)}$ is highest not $P{Y|T}$. If she knows thier gender then she should otimise to $P{Y|do(T),Z}$.

The causal network for this problem is shown below. In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men. This is what makes treatment $A$ appear better if you look at the conditional distribution. The fact that a patient took treatment $A$ tells you that they are more likely to be female. In the terminology defined in section \ref{}, gender is a confounding variable and we must adjust for it to obtain unbiased estimates of the causal effect of the treatment. As we saw from the examples in chapter \ref{XXX}. Determining when you really care about the interventional as opposed to the conditional distribution is a sublte but very important question that is frequently overlooked.

@@

Some of these variables my be latent or unobserved, in that we do not have measurements for them in the available data. However, their position in the network is assumed to be known. For example if we were trying to estimate ...

@@

\subsection*{The do calculus}
Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting?

The do calculus arises from d-seperation.

How do we take our causal frameworks and use them to map information from the observational to interventional settings?

@@

a) determine if the query can be translated into an expression involving only distributions over observed variables. In other words, determine if the query is identifiable given the assumptions encoded by the network

b) if it is identiable, return the required expression

\todo{insert link to working javascript version}

@@

\section*{Causal Discovery}
We now move to the much more general problem of learning a causal graph from observational data. Amazingly, it is possible to infer some aspects of causal structure with very general assumptions. The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others. In addition, there are subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause. These clues are the key to causal discovery algorithms, which attempt to learn causal structure from non-experimental data.

\paragraph*{You want to learn the graph}
In the causal discovery problem we make much much broader assumptions about the structure of the graph. Ie that it is acyclic. and may or may not have latent variables. We do not assume the existance or directions of any links between the variables.

What's more, we wish to do this from non-experimental data (although there is also work on infering causal structure from experimental data) REFERENCES and it would be natural to connect the two (ANY REFERENCES THERE?).

>>>>

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}

@@

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }

%\tikzset{
%    %Define standard arrow tip
%    >=stealth',
%    %Define style for boxes
%    punkt/.style={
%           circle,
%           rounded corners,
%           draw=black, thick,
%           text width=1em,
%           minimum height=1em,
%           text centered},
%    punktrect/.style={
%    rectangle, 
%    rounded corners, 
%    % fill=black!10,
%    draw=black, thick,
%    minimum height=1em, 
%    text centered},
%    % Define arrow style
%    pil/.style={
%           o->,
%           thick,
%           shorten <=2pt,
%           shorten >=2pt,}
%}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

There is a connection here to the degree to which the model is invariant. A model that is invariant to more things is more explainitory. We could look at levels of hierachy. There is causal graphical models, which are invarient to interventions on variables. Can I tie this to a physical system and give more indepth examples of the nested tier of hieriachies?

I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why.

Sometimes the causal component is obvious.

Relationship to generalization. Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time. The assumption is there are less places for change to come in.

If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.

The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)? This may be a reason for using structure learning type algorithms even if you are simply doing prediction.

\subsubsection*{The role of assumptions in causal inference}

Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.

Does predictive accuracy indicate a good causal model?

Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.

Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize). (and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)

Is causal inference possible (are the conclusions ever valid)?

Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.

The challenges and limitations of infering causal effects from observation data are numerous. However, there are problems we need to solve. The questions is are these problems so serious as to warrent discregarding all non-experimental data. I think answer is a clear no. Need ongoing validation, with assumptions. We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).

\subsection*{Relationship to interpretability}

A desire for interpretability indicates that something has been left out of the loss function.

Are causal/interpretable models more reliable or more likely to generalize well?

One form of interpretability gives people insight into what the features are that the model is relying on.

If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.

Specifically, people can
\begin{itemize}
\item rule out many possible features as highly unlikely to be relevent to a problem
\end{itemize}

People have access to a lot of detailed prior knowledge.

\subsection*{Relationship to transfer learning}

find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time). Causal models predict the outcome of actions. We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike. So we have to make representations such that things are stable.

This is tightly related to generalizability. If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results. We don't include , the color of the doctors shirt, the size of the room, ...

For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales. However, this relationship may be very unstable over time because the ad slots are sold at auction. The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword. However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.

@@

\paragraph*{There are two reasons why correlation is not causation}. The first is related to variance and over-fitting. Obversations are noisy. With a finite dataset with enough variables we will be able to find some that are completely unrelated but correlate purely by chance. \ref{figure - search for random correlation}. The second arises from bias, typically introced by an un-observed confounding variable. In this case, variables are correlated not by chance. We would expect the relationship to hold if we sampled more data. However, they are not causally related in that intervening to set one would not likely effect the other. EXAMPLE WITH FIGURE.

@@

Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to. Which of them require casual inference? How can we identify characterists of a prolem that make causal modeling important?

@@

The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.

Consider speech recognition. You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution). A similar arguament could be made for many applications of machine translation and image classification.

In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. We can be fairly confident that the process will not change. If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).

What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains. However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifing spam and automated essay marking systems are similar. The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label. Spammers and students will modify their writing in order to optimise thier results. The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided. (What would the nature of the features have to be such that change did not occur? - they would have to be causes of the label).

What about predicting the risk of death in patients with pnumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pnemonia. They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are;
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}

Demand prediction seems relatively straightforward. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However there is a potential data sensoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is another case with some subleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)

Having considered these examples we can now identify some general aspects of the problem that determine wheter or not we require a causal model.

\begin{itemize}
\item Does acting on the predictions of the model change the mapping from features to target? (at least if the decision process is open to scrutiny). In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
\item

 Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable. The mechanism is known but the function may be complex. Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models? Yes, if the way in which features get their values changes, then causal models will be invarient to that in a way that non-causal ones are not.

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and priortise those for modelling changes. 
\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its ok...
\end{itemize}

I hope these examples gave you a feel for the richness and subtlties of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

@@

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}). If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.

\begin{figure}[h]
\centering
\caption{General Bayesian network over 3 variables}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below left=of 1](2){$X_{2}$};
\node[main node, below right=of 1](3){$X_{3}$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (2) edge node {} (3);

\end{tikzpicture}
\end{figure}

@@

A causal network represents much more information than a bayesian network with identical structure. GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS. TAKE THE GRAPH IN WITH THREE NODES. ASSUME EVERYTHING IS BINARY AND. FIGURE OUT HOW MANY VARIABLES ARE NEEDED TO DESCRIBE ALL POSSIBLE INTERVENTIONS VS THE BAYESIAN NETWORK. This may be surprising since their conditional probability distributions are estiated from data in the same way. It is only possible due to the additional assumptions inherent in a causal network.

@@

\subsubsection*{Limitations of causal bayesian networks}
A number of critisims have been leveled at this way of modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-determinisitic ways \ref{causality in complex interventions,Cartwright 2007}. However, the deterministic $do()$ opperater turns out to be surprisingly effective at modelling more complex interventions. One simply has to add additional variables to the graph that model how our interventions may actually take effect. For example, our drug treatment case assumed we set the drug that each person recieves. But what if some people in fact fail to take the prescribed treatment. How can we model this within the framework of deterministic interventions? We simply add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually recieve. See figure XXXX

Insert figure of randomized treatment with imperfect compliance here.

What questions can't be phrased using this terminology
- certain counterfactual queries
- cyclic/dynamics
- unroll to deal with cycles

@@

@@

Once the data is broken down by gender, it appears that treament B is better for both men \emph{and} women. This is the paradox. If the doctor were unaware of the gender, it seems that she should advise treatment A. But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be.

How does causaul modelling reslove this paradox? Let's represent the treatment presribed by the variable $T$, the gender of the patient by $Z$ and wheter or not the treatment was successful by $Y$. The key to resolving the paradox is that the doctor really cares about the interventional distribution $P{Y|do(T)}$, not the standard conditional distributions $P{Y|T}$. The results from table XXX shows that $\P{Y|T=A}$ is higher than $\P{Y|T=B}$. If she has to suggest a treatment for a patient without knowing their gender, she should choose the treatment for which $P{Y|do(T)}$ is highest not $P{Y|T}$. If she knows thier gender then she should otimise to $P{Y|do(T),Z}$.

The causal network for this problem is shown in figure XXX. In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men. This is what makes treatment $A$ appear better if you look at the conditional distribution. The fact that a patient took treatment $A$ tells you that they are more likely to be female. In the terminology defined in section \ref{}, gender is a confounding variable and we must adjust for it to obtain unbiased estimates of the causal effect of the treatment. As we saw from the examples in chapter \ref{XXX}. Determining when you really care about the interventional as opposed to the conditional distribution is a sublte but very important question that is frequently overlooked.

@@

Some of these variables my be latent or unobserved, in that we do not have measurements for them in the available data. However, their position in the network is assumed to be known. For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtaine on graduating. Even if we have no data to directly measure peoples inherent ability, becuase it influences two of the variables were are modelling, we must include it in the graph.

@@

\subsection*{Mapping from observational to interventional distributions}
Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting to the interventional one. Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional idependencies between them.

\begin{theorem}
\textbf{(Local markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
\end{theorem}

The set of conditional independence relations given by the local markov condition can enforce additional independences that also hold in all distributions that are compatible with $G$. D-separation is an algorithm that extends the local markov property to find these additional independences. It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.

By leveraging conditional independencies we can effectivly localize the effect of an intervention to a specific part of a larger graph. This gives rise to the do-calculus \cite{Pearl2000}. The do calculus consists of three rules. They derive from the causal information encoded in a causal network and the properties of d-separation and do not require any addition assumptions other than that of specifying the causal network.

\subsubsection*{Rule 1}
This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention. The intervention $do(\boldsymbol{X}=\boldsymbol{x})$ changes a causal network, $G$, in a simple way. Variables in $\boldsymbol{X}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{x}$. This corresponds to deleting the edges with arrows into variables in $\boldsymbol{X}$  (see figure \ref{fig:mutilatednet}). The resulting 'mutilated' network $G_{\overline{\boldsymbol{X}}}$ remains a causal network and d-separation still applies.

If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
\begin{equation}
\label{eq:Do1}
 P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
\end{equation}

$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
\begin{equation}
\label{eq:Do12}
 P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
\end{equation}

This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.

\begin{figure}[h]
\centering
\caption{Intervention in a causal bayesian network}
\label{fig:mutilatednet}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{original network, $G$}
\label{fig:mutilatedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$W$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (2) edge node {} (3)
    (4) edge node {} (2);

\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
\label{fig:mutilatedAfter}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$x$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$W$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (3)
    (2) edge node {} (3);

\end{tikzpicture}
\end{subfigure}
\end{figure}

\subsubsection*{Rule 2}
Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$. You can think of this as when correlation is causation. It is easiest to understand by explicitly including the intervention process in the graphical model. We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on (figure \ref{fig:augmentedNetwork}). Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$. If $X = \epsilon$ the distribution of $X$ is what it was without intervention. Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$. We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.

if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:

\begin{equation}
\label{eq:Do2}
P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
\end{equation}

$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
\begin{equation}
\label{eq:Do22}
P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
\end{equation}

If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.

\begin{figure}[h]
\centering
\caption{Augmentated causal network with intervention}
\label{fig:augmentedNetwork}
\begin{subfigure}[t]{0.4\textwidth}
\centering
\caption{original network, $G$}
\label{fig:augmentedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left = of 1](4){$U_{1}$};
\node[main node, above right = of 1](5){$U_{2}$};

 \path[every node/.style={font=\sffamily\small}]
 	(2) edge node {} (3)
 	(4) edge node {} (2)
 		edge node {} (1)
 	(5) edge node {} (3)
 		edge node {} (1)
 	(1) edge node {} (2)
 		edge node {} (3);

\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.4\textwidth}
\centering
\caption{augmented network after the intervention $do(Z=z)$, $G^{\dagger}_{\overline{Z}}$}
\label{fig:augmentedAfter}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.3cm,
  thick,main node/.style={punkt}]

\node[main node](1){$z$};
\node[below left=of 1](2){};
\node[below right=of 1](3){};
\node[main node, below left=of 2](4){$X$};
\node[main node, below right=of 3](5){$Y$};
\node[above right=of 1](6){};
\node[above left=of 1](7){};
\node[main node, above left=of 7](8){$U_{1}$};
\node[main node, above right=of 6](9){$U_{2}$};
\node[main node, above left=of 4](10){$\hat{X}$};

 \path[every node/.style={font=\sffamily\small}]
 	(4) edge node {} (5)
 	(10) edge node {} (4)
 	(8) edge node {} (4)
 	(9) edge node {} (5)
 	(1) edge node {} (4)
 		edge node {} (5);

\end{tikzpicture}
\end{subfigure}
\end{figure}

\subsubsection*{Rule 3}
This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.

if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
\begin{equation}
\label{eq:Do3}
P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
\end{equation}
$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
\begin{equation}
\label{eq:Do32}
P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
\end{equation}
The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$. This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.

@@

\begin{enumerate}
\item determine if the query can be translated into an expression involving only distributions over observed variables. In other words, determine if the query is identifiable given the assumptions encoded by the network
\item if it is identiable, return the required expression 
\end{enumerate}

\todo{insert link to working javascript version}
\todo{figure with typical examples}

@@

\section*{Causal Discovery}
We now move to the much more general problem of learning a causal graph from observational data. In this setting we make much broader assumptions about the structure of the graph. For example, that it is acyclic or that we have no unmeasured confounding variables. We do not assume the existance or directions of any links between the variables. Amazingly, it is possible to infer some aspects of causal structure with such general assumptions. The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others. In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause. These clues are the key to causal discovery algorithms.

There is also work on infering causal structure from experiental data \ref{}. We discuss this further in section XXX.
